---
title: "final_project_Juan_ML"
format: html
editor: visual
---

Modelling

# Ⅰ. Set up

## 1. Install packages

```{r}
#| message: false
#| warning: false

.libPaths(c("/home/jg44526/R_lib_4.4.2/", .libPaths()))

library(tidymodels)   
library(finetune)     
library(vip)          
library(xgboost)      
library(tidyverse)    
library(doParallel)   
library(lme4)        
library(here)         
library(quarto) 

here::here()
```

## 2. Load data

```{r}
corn_data <- read_csv("../data/train_data.csv") %>%
  mutate(year = as.factor(year))

#glimpse(corn_data)
```

# II. XGBoost

## 1. Pre-processing

```{r}
set.seed(123)
corn_data_split <- initial_split(
  corn_data, 
  prop = .7, # proption of split same as previous codes
  strata = yield_mg_ha_mean  # Stratify by target variable
  )

corn_data_split
```

### 1.1 Data split

```{r}
set.seed(123)
corn_train <- training(corn_data_split)  # 70% of data
corn_train #This is your training data frame

corn_test <- testing(corn_data_split)    # 30% of data
corn_test
```

### 1.2 Check target distribution

```{r distribution}
density_plot <- ggplot() +
  geom_density(
    data = corn_train,
    aes(x = yield_mg_ha_mean),
    color = "red"
  ) +
  geom_density(
    data = corn_test,
    aes(x = yield_mg_ha_mean),
    color = "blue"
  ) +
  labs(
    title = "Distribution of yield in train vs test",
    x = "yield_mg_ha_mean",
    y = "Density"
  )

density_plot

cat("Saving...... density plot of train and test data\n")

# Optional save
# ggsave(
#   plot = density_plot,
#   path = here("output", "png"),
#   filename = "density_plot_test_train.png",
#   height = 6,
#   width = 9,
#   dpi = 600
# )
```

### 1.3 Data processing with recipe

```{r Check NAs}
library(tibble)

na_summary <- tibble(
  column = names(corn_train),
  na_ratio = colMeans(is.na(corn_train))
) %>%
  arrange(desc(na_ratio))

na_summary
```

```{r}
cat("\nCreate recipe for data preprocessing\n")

corn_recipe <- recipe(yield_mg_ha_mean ~ ., data = corn_train) %>%
  step_rm(site, year) %>%
  step_unknown(previous_crop) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_indicate_na(all_of(c("om_pct", "soil_ph", "soilk_ppm", "soilp_ppm"))) %>%
  step_nzv(all_predictors())

# Prep the recipe to estimate any required statistics
corn_prep <- corn_recipe %>% 
  prep()

# Examine preprocessing steps
corn_prep

```

## 2. Training

### 2.1 Model specification

```{r xgb_spec}
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  min_n = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  sample_size = tune(),
  mtry = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_spec
```

### 2.2 Resampling setup

```{r}
set.seed(123)

# Create 10-fold cross-validation resampling object from training data
 resampling_foldcv <- vfold_cv(corn_train, 
                               v = 20,
                               strata = yield_mg_ha_mean)
 
# Create leave one year out cross-validation (cv) object from the sampling data
# resampling_fold_loyo <- group_vfold_cv(corn_train,
#                                        group = year)
 
# Create leave one location out cv object from the sampling data
# resampling_fold_loso <- group_vfold_cv(corn_train,
#                                        group = site)
```

### 2.3 Hyperparameter grid with Latin Hypercube Sampling

```{r }
# Use prepped predictor data so mtry range matches processed columns better
predictor_data <- bake(corn_prep, new_data = NULL) %>%
  dplyr::select(-yield_mg_ha_mean)

xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  loss_reduction(),
  sample_prop(),
  finalize(mtry(), predictor_data),
  size = 30
)

xgb_grid
```

```{r}
ggplot(
  data = xgb_grid,
  aes(x = tree_depth, y = min_n)
) +
  geom_point(
    aes(color = factor(learn_rate), size = trees),
    alpha = 0.5,
    show.legend = FALSE
  )
```

```{r Workflow}
xgb_wf <- workflow() %>%
  add_recipe(corn_recipe) %>%
  add_model(xgb_spec)

xgb_wf
```

## 3. Detecting total cores

-   This help us to detect the total number of available cores and then do a parallel processing on each of them

```{r}
# Detect Cores for Slurm (Cluster - Sapelo2)

n_cores <- as.numeric(Sys.getenv("SLURM_CPUS_ON_NODE"))
if (is.na(n_cores)) n_cores <- max(parallel::detectCores() - 1, 1)

cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)

cat("Found and registered ", n_cores, " cores to work with\n")

```

## 4. Model Tuning

```{r xgb_grid_result}
set.seed(123)

xgb_res <- tune_race_anova(
  object = xgb_wf,
  resamples = resampling_foldcv,
  grid = xgb_grid,
  metrics = metric_set(rmse, rsq),
  control = control_race(
    save_pred = TRUE,
    verbose_elim = TRUE
  )
)

# Stop cluster cleanly
stopCluster(cl)
registerDoSEQ()

```

Let's check the race results

```{r}
plot_race(xgb_res)
collect_metrics(xgb_res)
#saveRDS(xgb_res,"xgb_res.rds")
xgb_res <- read_rds("/home/jg44526/2026_final_project/code/output_juan_modeling_1/xgb_res_juan_modeling.rds")
```

## 5. Select Best Models

```{r}
best_rmse <- select_best(xgb_res, metric = "rmse") %>%
  mutate(source = "best_rmse")

best_rmse_pct_loss <- select_by_pct_loss(
  xgb_res,
  min_n,
  metric = "rmse",
  limit = 1
) %>%
  mutate(source = "best_rmse_pct_loss")

best_rsq <- select_best(xgb_res, metric = "rsq") %>%
  mutate(source = "best_rsq")

best_rsq_pct_loss <- select_by_pct_loss(
  xgb_res,
  min_n,
  metric = "rsq",
  limit = 1
) %>%
  mutate(source = "best_rsq_pct_loss")

bind_rows(
  best_rmse,
  best_rmse_pct_loss,
  best_rsq,
  best_rsq_pct_loss
) %>%
  dplyr::select(source, everything())

# Use RMSE winner as final choice
best_params <- best_rmse %>%
  dplyr::select(-source)

best_params

```

## 6. Compare and Finalize Model

```{r Finalized workflow}
final_wf <- finalize_workflow(xgb_wf, best_params)

final_wf
```

## 7. Resample final model

```{r final_spec_fit}
res_cv_final <- fit_resamples(
  final_wf,
  resamples = resampling_foldcv,
  metrics = metric_set(rmse, rsq)
)
# # res_loyo_final <- fit_resamples(
#   final_wf,
#   resamples = resampling_fold_loyo,
#   metrics = metric_set(rmse, rsq)
# )
# 
# # res_loso_final <- fit_resamples(
#   final_wf,
#   resamples = resampling_fold_loso,
#   metrics = metric_set(rmse, rsq)
# )
# saveRDS(res_cv_final, "res_cv_final.rds")
res_cv_final<-read_rds("/home/jg44526/2026_final_project/code/output_juan_modeling_1/xgb_res_juan_modeling.rds")
collect_metrics(res_cv_final)
# collect_metrics(res_loyo_final)
# collect_metrics(res_loso_final)
```

## 8. Final Fit and Predictions

## Final validation on held-out test set

```{r final_fit}

set.seed(123)

final_fit <- last_fit(final_wf, split = corn_data_split)

saveRDS(final_fit,"./output_juan_modeling_1/final_fit.rds")

final_preds <- final_fit %>%
  collect_predictions()

final_preds

final_metrics <- final_fit %>%
  collect_metrics()

final_metrics
```

## 8. Fit on training set

```{r}
train_fit <- fit(final_wf, data = corn_train)

train_pred <- augment(train_fit, new_data = corn_train)

train_metrics <- metrics(
  train_pred,
  truth = yield_mg_ha_mean,
  estimate = .pred
)

train_metrics

```

## 9. Predicted vs Observed Plot

```{r}
final_fit_plot <- final_preds %>%
  ggplot(aes(x = yield_mg_ha_mean, y = .pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  geom_smooth(method = "lm", se = FALSE) +
  coord_cartesian(xlim = c(1, 25), ylim = c(1, 25)) +
  labs(
    title = "Predicted vs Observed",
    x = "Observed yield",
    y = "Predicted yield"
  )

final_fit_plot

# Optional save
 ggsave(
  plot = final_fit_plot,
   path = here("output", "png"),
   filename = "final_fit_plot_juan_model.png",
   height = 6,
   width = 9,
   dpi = 300
 )
```

## 10. Variable Importance

```{r final_spec}
# Extract fitted parsnip model first
xgb_fit_obj <- extract_fit_parsnip(train_fit)

vi_tbl <- vip::vi(xgb_fit_obj)

saveRDS(vi_tbl, "./output_juan_modeling_1/vip.rds")

# Full importance plot
vi_tbl %>%
  mutate(Variable = forcats::fct_reorder(Variable, Importance)) %>%
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL, title = "Variable importance")

# Top 20 importance plot
final_vi_plot <- vi_tbl %>%
  slice_max(order_by = Importance, n = 20) %>%
  mutate(
    Variable = stringr::str_trunc(Variable, width = 25),
    Variable = forcats::fct_reorder(Variable, Importance)
  ) %>%
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(
    y = NULL,
    title = "Top 20 variable importance"
  ) +
  theme(axis.text.y = element_text(size = 8))

final_vi_plot

# Optional save
ggsave(
   plot = final_vi_plot,
   path = here("./output_juan_modeling_1/vi", "png"),
   filename = "final_spec_plot.png",
   height = 6,
   width = 9,
   dpi = 600
 )
```

```{r Save fitted final model}
saveRDS(train_fit, "final_model.rds")
```

# 11. Predict external test_data

```{r Build submission file}
test_submission <- read_csv("../data/testing_submission.csv")

test_out <- bind_cols(test_submission, pred_test) %>%
  dplyr::select(-any_of("yield_mg_ha_mean")) %>%
  rename(yield_mg_ha = .pred)

write_csv(test_out, "../data/testing_submission_finished.csv")

test_out
```

#Ming Revised \## Do test data Let's predict the test data

```{r save model}
# 1. Load real test data
real_test <- read_csv("../data/test_data.csv") %>%
  mutate(year = as.factor(year))

# 2. Fit final model on ALL available training data
# do not use corn_train，use full data: corn_data
final_model <- fit(final_wf, data = corn_data)

# 3. Predict real test data
real_test_pred <- predict(final_model, new_data = real_test) %>%
  bind_cols(real_test)

real_test_pred
```

```{r save}
write_csv(
  real_test_pred,
  "./output_juan_modeling_1/real_test_predictions.csv"
)
```
