---
title: "final_project"
format: html
editor: visual
---

# Data wrangling

## Training data

### Ⅰ. Growing data

#### 1. Set up

##### 1.1 Install packages

```{r}
#| message: false
#| warning: false

#install.packages("nasapower")
#install.packages("beepr")
#install.packages("sf")
#install.packages("rnaturalearthdata")
#install.packages("daymetr")
#install.packages("remotes")
#remotes::install_github("ropensci/USAboundaries")
#remotes::install_github("ropensci/USAboundariesData")
#install.packages("soilDB")
#install.packages("aqp")

library(soilDB) # for pulling data from SoilGrid
library(aqp)
library(tidyverse)
library(tidyr)
library(janitor)
library(tidymodels)
library(xgboost)
library(purrr)
library(stringr)
library(lubridate)
library(nasapower)
library(beepr)
library(USAboundaries)
library(sf)
library(rnaturalearth)
library(daymetr)
```

##### 1.2 Import taining data

```{r}
train_meta <- read_csv("../data/training_meta.csv")
train_soil <- read_csv("../data/training_soil.csv")
train_trait <- read_csv("../data/training_trait.csv")

#glimpse(train_meta)
#glimpse(train_soil)
#glimpse(train_trait)
```

#### 2. EDA

##### 2.1 Summary

```{r}
summary(train_meta)
summary(train_soil)
summary(train_trait)
```

##### 2.2 Distinct year, site, hybrid

```{r}
train_meta %>% distinct(year)
train_soil %>% distinct(year)
train_trait %>% distinct(year)

train_meta %>% distinct(site)
train_soil %>% distinct(site)
train_trait %>% distinct(site)

train_trait %>% distinct(hybrid)
```

There is 5,027 hybrids.

##### 2.3 Check non-US site

```{r}
# check if all sites are in the US.

# read US boundary
usa <- ne_countries(
  scale = "medium",
  country = "United States of America",
  returnclass = "sf"
)

# convert points to sf objects
points <- train_meta %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# check whether points are within the US
inside <- st_within(points, usa)

# label results (clearer)
train_meta_check <- train_meta %>%
  mutate(in_us = lengths(inside) > 0)

# view points not in the US
not_in_us <- train_meta_check %>%
  filter(!in_us)

not_in_us
```

According to the latitude and longitude, GEH1, ONH1, and ONH2 are located outside the United States and would normally be excluded based on the assignment requirement to include only U.S. data. However, since the testing submission file includes the ONH3 site, which is located in Canada, ONH1 and ONH2 are retained in the training dataset for consistency.

##### 2.4 Check plant and harvest timeline

```{r}
# check site(s) in train_data but not in test_data
test_data <- read_csv("../output/test_data.csv")
train_data <- read_csv("../output/train_data.csv")
setdiff(unique(test_data$site), unique(train_data$site))

# check plant and harvest date
train_trait %>%
  distinct(site, date_planted, date_harvested) %>%
  arrange(site, date_planted)

# check plant and harvest month
site_window <- train_trait %>%
  mutate(
    date_planted   = mdy(date_planted),
    date_harvested = mdy(date_harvested),
    plant_month    = month(date_planted),
    harvest_month  = month(date_harvested)
  ) %>%
  group_by(site) %>%
  summarise(
    plant_start   = floor(quantile(plant_month, 0.25, na.rm = TRUE)),
    plant_end     = ceiling(quantile(plant_month, 0.75, na.rm = TRUE)),
    harvest_start = floor(quantile(harvest_month, 0.25, na.rm = TRUE)),
    harvest_end   = ceiling(quantile(harvest_month, 0.75, na.rm = TRUE)),
    n_plant_na    = sum(is.na(plant_month)),
    n_harvest_na  = sum(is.na(harvest_month)),
    .groups = "drop"
  )

site_window
```

##### 2.5 Check A/B and B/A in hybrid

```{r}
hyb_check <- train_trait %>%
  distinct(hybrid) %>%
  separate(hybrid, into = c("p1", "p2"), sep = "/", remove = FALSE) %>%
  mutate(rev = paste(p2, p1, sep = "/"))

sum(hyb_check$rev %in% hyb_check$hybrid)

test_trait_clean <- read_csv("../output/test_trait_clean.csv")
train_trait_clean <- read_csv("../output/train_trait_clean.csv")

# test: how many new hybrid
 length(setdiff(unique(test_trait_clean$hybrid), unique(train_trait_clean$hybrid)))

# overlap rate
 length(intersect(unique(test_trait_clean$hybrid), unique(train_trait_clean$hybrid))) /
 length(unique(test_trait_clean$hybrid))
```

There are 6 pairs that contain A/B and B/A in hybrid.

Only 0.09783631 hybrid overlaps in train and test dataset.

#### 3. Data cleaning

##### 3.1 Clean train_soil

```{r}
train_soil_clean <- read_csv("../data/training_soil.csv") %>%
  clean_names() %>%
  rename(soil_ph = soilp_h) %>%
  
  # delete all duplicate rows
  distinct() %>%
  
  # Remove year from site
  mutate(
    site_no_year = str_remove(site, "_\\d{4}$"),
    site_base = str_remove(site_no_year, "_\\d+$")
  ) %>%
  
  # Create a soil profile to check whether data are identical, within the same year and base site
  mutate(
    soil_profile = paste(soil_ph, om_pct, soilk_ppm, soilp_ppm, sep = " | ")
  ) %>%
  
  # Within the same year and base site, check whether different replicates are identical
  group_by(year, site_base) %>%
  mutate(
    n_profile = n_distinct(soil_profile),
    site_final = if_else(n_profile == 1, site_base, site_no_year)
  ) %>%
  ungroup() %>%
  
  # Remove duplicated records again
  distinct(year, site_final, soil_ph, om_pct, soilk_ppm, soilp_ppm, .keep_all = TRUE) %>%
  
  # Output final columns
  transmute(
    year,
    site = site_final,
    soil_ph,
    om_pct,
    soilk_ppm,
    soilp_ppm
  ) %>%
  
  # remove non-US site
  filter(site != "GEH1") %>%
  
  arrange(desc(year), site)

write_csv(train_soil_clean, "../output/train_soil_clean.csv")
#train_soil_clean %>% filter(str_detect(site, "\\d{4}$"))
#train_soil_clean %>% filter(str_detect(site, "^MOH1"))
```

##### 3.2 Clean train_trait

```{r}
train_trait_clean <- read_csv("../data/training_trait.csv") %>%
  clean_names() %>%
  
  mutate(
    site = str_replace(site, "^([A-Z0-9]+)[abc]$", "\\1"),
    site = case_when(
      site %in% c("TXH1-Dry", "TXH1-Early", "TXH1-Late") ~ "TXH1",
      TRUE ~ site
    ),
    
    date_planted = mdy(date_planted),
    date_harvested = mdy(date_harvested),
    
    planted_month = month(date_planted, label = TRUE, abbr = TRUE),
    harvested_month = month(date_harvested, label = TRUE, abbr = TRUE),
    
#    growing_days = as.numeric(date_harvested - date_planted),
#    planted_doy = yday(date_planted),
#    harvested_doy = yday(date_harvested),
  ) %>%
  
   separate(hybrid, into = c("parent1", "parent2"), sep = "/", remove = FALSE) %>%
  
#  dplyr::select(
#    year, site, 
#    hybrid, parent1, parent2, 
#    planted_month, harvested_month,
#    yield_mg_ha # No grain_moisture info in the predict-dataset + it seems highly related to yield ==> it is excluded.
#  ) %>%
  
  # remove non-US site
  filter(site != "GEH1") %>%
  
    # aggregate replicate and block
  group_by(year, site, hybrid) %>%
  summarise(
    parent1 = first(parent1), 
    parent2 = first(parent2), 
    planted_month = first(planted_month),
    harvested_month = first(harvested_month),
    yield_mg_ha_mean = mean(yield_mg_ha, na.rm = TRUE),
#    n_rep = n(),
#    yield_sd = sd(yield_mg_ha, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(yield_mg_ha_mean = round(yield_mg_ha_mean, 2))%>%
  
  arrange(desc(year), site, hybrid)

#train_trait %>%
#  group_by(year, site, hybrid) %>%
#  summarise(sd = sd(yield_mg_ha, na.rm = TRUE)) %>%
#  summarise(mean_sd = mean(sd, na.rm = TRUE))

#train_trait_clean %>%
#  group_by(hybrid) %>%
#  summarise(sd_between = sd(yield_mg_ha, na.rm = TRUE))

write_csv(train_trait_clean, "../output/train_trait_clean.csv")
```

##### 3.3 Clean train_meta

```{r}
train_meta_clean <- read_csv("../data/training_meta.csv") %>%
  clean_names() %>%
  
  mutate(
    # lowercase
    previous_crop = tolower(previous_crop),
    # clean crop
    previous_crop = case_when(
      is.na(previous_crop) ~ NA_character_,
      str_detect(previous_crop, "soy") ~ "soybean",
      str_detect(previous_crop, "corn") ~ "corn",
      str_detect(previous_crop, "wheat") ~ "wheat",
      str_detect(previous_crop, "cotton") ~ "cotton",
      str_detect(previous_crop, "sorghum") ~ "sorghum",
      str_detect(previous_crop, "lima") ~ "lima bean",
      str_detect(previous_crop, "cereal") ~ "cereal rye",
      str_detect(previous_crop, "fallow") ~ "fallow",
      str_detect(previous_crop, "peanut") ~ "peanut",
      str_detect(previous_crop, "beet") ~ "sugar beet",
      TRUE ~ "other"
    ),
    # combine rows with repeated sites
    site = case_when(
      site %in% c("TXH1- Dry", "TXH1- Early", "TXH1- Late") ~ "TXH1", 
      site %in% c("MOH1- rep 1", "MOH1- rep 2") ~ "MOH1",
      TRUE ~ site
    )
  ) %>%
 
  # remove repeated rows
  distinct(year, site, previous_crop, longitude, latitude, .keep_all = TRUE) %>%
  # average location info within repeated sites
  group_by(year, site, previous_crop) %>%
  summarise(
    longitude = mean(longitude, na.rm = TRUE),
    latitude = mean(latitude, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  
  # remove non-US site
  filter(site != "GEH1") %>%
  
  arrange(desc(year), site)

write_csv(train_meta_clean, "../output/train_meta_clean.csv")
#train_meta_clean %>% count(previous_crop, sort = TRUE)
#train_meta_clean %>% count(site, year) %>% filter(n > 1)
```

### Ⅱ. Weather data

#### 1. Import data from Daymet

##### 1.1 Find complete site, year info

```{r}
# Pull site-year info from train_trait_clean
full_site_year <- train_trait_clean %>%
  distinct(site, year)

full_site_year

# Pull site-year-specific coordinates from train_meta_clean
site_coord <- train_meta_clean %>%
  select(site, year, latitude, longitude) %>%
  distinct(site, year, .keep_all = TRUE)

site_coord

# Merge by both site and year
full_site_year_coord <- full_site_year %>%
  left_join(site_coord, by = c("site", "year"))

full_site_year_coord
```

```{r}
# missing site-year in trait but not in meta
missing_site_year <- full_site_year %>%
  anti_join(site_coord, by = c("site", "year"))

missing_site_year
```

```{r}
missing_site_coord_candidates <- train_meta_clean %>%
  semi_join(missing_site_year %>% distinct(site), by = "site") %>%
  select(site, year, latitude, longitude) %>%
  arrange(site, year)

missing_site_coord_candidates
```

```{r}
# Based on site not year
site_mean_coord <- train_meta_clean %>%
  group_by(site) %>%
  summarise(
    lat_mean = mean(latitude, na.rm = TRUE),
    lon_mean = mean(longitude, na.rm = TRUE),
    .groups = "drop"
  )

# Merge and fill in NAs
full_site_year_coord_filled <- full_site_year_coord %>%
  left_join(site_mean_coord, by = "site") %>%
  mutate(
    latitude  = if_else(is.na(latitude),  lat_mean, latitude),
    longitude = if_else(is.na(longitude), lon_mean, longitude)
  ) %>%
  select(site, year, latitude, longitude)

full_site_year_coord_filled

# Check if there is still NA or not
full_site_year_coord_filled %>% filter(is.na(latitude) | is.na(longitude))
```

```{r}
# Check all TXH coordinate info
train_meta_clean %>%
  filter(str_detect(site, "TXH")) %>%
  arrange(site, year)
```

```{r}
# Create TXH4 coordinate
txh_coord <- train_meta_clean %>%
  filter(str_detect(site, "TXH")) %>%
  filter(longitude > -100) %>%  # remove 2014 TXH2 (seems like outlier)
  summarise(
    lat = mean(latitude),
    lon = mean(longitude)
  )

txh_coord
```

```{r}
full_site_year_coord_complete <- full_site_year_coord_filled %>%
  mutate(
    latitude  = if_else(site == "TXH4", txh_coord$lat, latitude),
    longitude = if_else(site == "TXH4", txh_coord$lon, longitude)
  )

full_site_year_coord_complete
```

```{r}
train_meta_clean %>%
  group_by(site, year) %>%
  summarise(n_crop = n_distinct(previous_crop)) %>%
  filter(n_crop > 1)
```

```{r}
meta_crop <- train_meta_clean %>%
  select(site, year, previous_crop) %>%
  distinct(site, year, .keep_all = TRUE)
meta_crop

full_data <- full_site_year_coord_complete %>%
  left_join(meta_crop, by = c("site", "year"))
full_data
```

##### 1.2 Pull weather data

```{r}
# help("download_daymet")

# full_data_daym <- full_data %>%
#  mutate(
#    weather = pmap(
#      list(year, site, latitude, longitude),
#     function(.y, .site, .lat, .lon) {
#        download_daymet(
#         site = .site,
        #   lat = .lat,
        #   lon = .lon,
        #   start = .y,
        #   end = .y,
        #   simplify = TRUE,
        #   silent = TRUE
        # ) %>%
        #   rename(
        #     weather_year = year,
        #     weather_site = site,
        #     weather_latitude = latitude,
        #     weather_longitude = longitude
 #        #   )
 #      }
 #    )
 #  ) %>%
 #  unnest(weather) %>%
 #  pivot_wider(names_from = measurement,
 #              values_from = value
 #              ) %>%
 #  janitor::clean_names()
 # 
 # full_data_daym

 # write_csv(full_data_daym, "../data/full_data_daym.csv")

weather_tain <- read_csv("../data/full_data_daym.csv")
```

-   Variables included:
    -   day length (secs/day)\
    -   precipitation (mm/day)\
    -   shortwave radiation (W/m2)\
    -   snow water equivalent (kg/m2)\
    -   maximum air temperature (C)\
    -   minimum air temperature (C)\
    -   water vapor pressure (Pa)

#### 2. EDA

```{r}
# overview
summary(weather_tain)

# density plot to see distribution
weather_tain %>%
  pivot_longer(cols = dayl_s:vp_pa) %>%
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~name, scales = "free")
```

#### 3. Feature engineering

##### 3.1 Fix grow month

Since planting and harvest date information is not available in the testing dataset, it is difficult to directly align weather data with the actual growing season. To address this, I summarize the range of planting and harvest months for each site using the training data. Specifically, I use the minimum planting month as the start of the growing period and the maximum harvest month as the end. These site-specific month ranges are then used to extract and aggregate weather data for both the training and testing datasets. I assume that this time period reasonably represents the growing season for corn in the testing data.

```{r}
month_levels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

grow_month <- train_trait_clean %>%
  mutate(
    planted_month   = str_trim(str_to_title(planted_month)),
    harvested_month = str_trim(str_to_title(harvested_month)),
    plant_mon_num   = match(planted_month, month_levels),
    harvest_mon_num = match(harvested_month, month_levels)
  ) %>%
  group_by(site) %>%
  summarise(
    start_mon = month_levels[min(plant_mon_num, na.rm = TRUE)],
    end_mon   = month_levels[max(harvest_mon_num, na.rm = TRUE)],
    .groups = "drop"
  )

grow_month

write_csv(grow_month, "../output/grow_month.csv")
```

##### 3.2 Merge grow month with weather data

```{r}
weather_tain_grow <- weather_tain %>%
  left_join(grow_month, by = "site") %>%
  dplyr::select(year, site, previous_crop, start_mon, end_mon, everything())

weather_tain_grow

setdiff(unique(weather_tain_grow$site), unique(grow_month$site))
```

##### 3.3 Create month column for weather info

```{r}
weather_tain_month <- weather_tain_grow %>%
  # Selecting needed variables
  dplyr::select(year, site, previous_crop, 
                start_mon, end_mon,
                latitude, longitude,
                yday,
                dayl.s = dayl_s, 
                prcp.mm = prcp_mm_day,
                srad.wm2 = srad_w_m_2,
                tmax.c = tmax_deg_c, 
                tmin.c = tmin_deg_c,
                vp.pa = vp_pa
                ) %>%
  # Creating a date class variable
  mutate(date_chr = paste0(year,"/",yday)) %>% 
  mutate(date = as.Date(date_chr, "%Y/%j")) %>%
  # Extracting month from date  
  mutate(month = month(date)) %>%
  mutate(month_abb = month(date, label = T)) %>%
  # calculate GDD
  mutate(
    tmean = (tmax.c + tmin.c) / 2,
    gdd = pmax(tmean - 10, 0) # T-base: 10 for corn https://fieldreport.caes.uga.edu/publications/C1320/growing-degree-days-for-sweet-corn-in-southern-georgia/
    )

weather_tain_month
```

##### 3.4 Cut growing season into 4 stages

```{r}
month_levels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

weather_stage <- weather_tain_month %>%
  mutate(
    month_abb = factor(as.character(month_abb), levels = month_levels, ordered = TRUE),
    start_mon = factor(start_mon, levels = month_levels, ordered = TRUE),
    end_mon   = factor(end_mon, levels = month_levels, ordered = TRUE)
  ) %>%
  filter(month_abb >= start_mon & month_abb <= end_mon) %>%
  group_by(year, site) %>%
  mutate(stage = ntile(yday, 4)) %>%
  ungroup() 

weather_stage
```

##### 3.5 Summarize daily weather variables based on growing stage

```{r}
weather_tain_month_sum <- weather_stage %>%
    mutate(
    month_abb = factor(as.character(month_abb), levels = month_levels, ordered = TRUE),
    start_mon = factor(start_mon, levels = month_levels, ordered = TRUE),
    end_mon   = factor(end_mon, levels = month_levels, ordered = TRUE)
  ) %>%
  filter(month_abb >= start_mon & month_abb <= end_mon) %>% 
  group_by(year, site, previous_crop, stage) %>%
  summarise(across(.cols = c(dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa
                             ),
                   .fns = mean,
                   .names = "mean_{.col}"
                   ),
            across(.cols = c(prcp.mm, gdd),
                   .fns = sum,
                   .names = "sum_{.col}"
                   )
            ) %>%
  ungroup()

weather_tain_month_sum
```

##### 3.6 Verify the calculation

```{r}
weather_stage %>%
  filter(year == 2014 & 
           site == "DEH1" &
           previous_crop == "soybean" &
           stage == 1) %>%
  summarise(tmax.c = mean(tmax.c),
            prcp.mm = sum(prcp.mm))
```

##### 3.7 Include stage in the column name

```{r}
weather_tain_month_sum_wide <- weather_tain_month_sum %>%
  pivot_longer(mean_dayl.s:sum_gdd) %>%
  mutate(varname = paste0(name, "_stage",stage)) %>%
  dplyr::select(-name, -stage) %>%
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(where(is.numeric), 
                ~round(.,2)
                )
         ) 

weather_tain_month_sum_wide

write_csv(weather_tain_month_sum_wide, "../output/weather_tain_month_sum_wide.csv")
```

### Ⅲ. Soil data

#### 1. Create full soil info

```{r}
train_soil_full <- full_data %>%
  select(-previous_crop) %>%
  left_join(train_soil_clean, by = c("site", "year"))

train_soil_full
```

#### 2. Check the missing info

```{r}
# area
train_soil_full %>%
  summarise(
    om_pct_na = sum(is.na(om_pct)),
    soil_ph_na = sum(is.na(soil_ph)),
    soilk_ppm_na = sum(is.na(soilk_ppm)),
    soilp_ppm_na = sum(is.na(soilp_ppm))
  )

# line
train_soil_full %>%
  filter(
    is.na(om_pct) |
    is.na(soil_ph) |
    is.na(soilk_ppm) |
    is.na(soilp_ppm)
  )
```

#### 3. Keep only the site-years that were pulled

```{r}
train_soil_needpull <- train_soil_full %>%
  filter(
    is.na(om_pct) |
    is.na(soil_ph) |
    is.na(soilk_ppm) |
    is.na(soilp_ppm)
  ) %>% distinct(site, latitude, longitude)

train_soil_needpull
```

#### 4. Pull data from SoilGrid

```{r}
glimpse(train_soil_needpull)

summary(train_soil_needpull$latitude)
summary(train_soil_needpull$longitude)

train_soil_needpull %>% 
  filter(is.na(latitude) | is.na(longitude))
```

```{r}
# # SoilGrids pull
#  sg_input_train <- train_soil_needpull %>%
#    filter(!is.na(latitude), !is.na(longitude)) %>%
#    distinct(latitude, longitude) %>%
#    mutate(id = paste0("pt_", row_number())) %>%
#    select(id, latitude, longitude)
# 
#  soilgrid_raw_train <- fetchSoilGrids(
#    x = sg_input_train,
#    loc.names = c("id", "latitude", "longitude"),
#    variables = c("phh2o", "soc"),
#    depth_intervals = "0-5"
#  )

```

#### 5. Organize data

```{r}
# names(soilgrid_raw_train)
```

```{r}
#h <- horizons(soilgrid_raw_train) %>% as.data.frame()
# s <- site(soilgrid_raw_train) %>% as.data.frame()
# 
# names(s)
# names(h)
```

```{r}
# Find column name
# soilgrid_fill_train <- s %>%
#   left_join(
#     h %>%
#       select(id, phh2omean, socmean),
#     by = "id"
#   ) %>%
#   transmute(
#     latitude,
#     longitude,
#     soil_ph_sg = phh2omean,
#     om_pct_sg = socmean * 0.1724
#   )
# 
# soilgrid_fill_train 
# 
# write_csv(soilgrid_fill_train, "../output/soilgrid_fill_train.csv")
```

#### 6. Merge back

```{r}
soilgrid_fill_train <- read_csv("../output/soilgrid_fill_train.csv")

train_soil_filled <- train_soil_full %>%
  left_join(soilgrid_fill_train, by = c("latitude", "longitude")) %>%
  mutate(
    soil_ph = coalesce(soil_ph, soil_ph_sg),
    om_pct  = coalesce(om_pct, om_pct_sg)
  ) %>%
  select(-soil_ph_sg, -om_pct_sg, -latitude, -longitude)

train_soil_filled
```

```{r}
# check NAs
train_soil_filled %>%
  summarise(
    soil_ph_na = sum(is.na(soil_ph)),
    om_pct_na  = sum(is.na(om_pct)),
    soilk_ppm_na = sum(is.na(soilk_ppm)),
    soilp_ppm_na = sum(is.na(soilp_ppm))
  )
```

```{r}
# label ones with SoilGrid data

# train_soil_filled <- train_soil_full %>%
#   left_join(soilgrid_fill, by = c("latitude", "longitude")) %>%
#   mutate(
#     soil_ph_from_sg = as.integer(is.na(soil_ph) & !is.na(soil_ph_sg)),
#     om_pct_from_sg  = as.integer(is.na(om_pct)  & !is.na(om_pct_sg)),
#     soil_ph = coalesce(soil_ph, soil_ph_sg),
#     om_pct  = coalesce(om_pct, om_pct_sg)
#   ) %>%
#   select(-soil_ph_sg, -om_pct_sg)
# 
# train_soil_filled
```

```{r}
# remove the number after site. use the mean value to fill calculated from the same base site 
# train_soil_filled2 <- train_soil_filled %>%
#   mutate(
#     site_base = str_remove(site, "\\d+$")
#   ) %>%
#   group_by(site_base) %>%
#   mutate(
#     soilk_base_mean = mean(soilk_ppm, na.rm = TRUE),
#     soilp_base_mean = mean(soilp_ppm, na.rm = TRUE)
#   ) %>%
#   ungroup() %>%
#   mutate(
#     soilk_ppm = if_else(
#       is.na(soilk_ppm) & is.finite(soilk_base_mean),
#       soilk_base_mean,
#       soilk_ppm
#     ),
#     soilp_ppm = if_else(
#       is.na(soilp_ppm) & is.finite(soilp_base_mean),
#       soilp_base_mean,
#       soilp_ppm
#     )
#   ) %>%
#   select(-site_base, -soilk_base_mean, -soilp_base_mean)
# 
# train_soil_filled2
```

```{r}
# train_soil_filled2 %>%
#   summarise(
#     soilk_ppm_na = sum(is.na(soilk_ppm)),
#     soilp_ppm_na = sum(is.na(soilp_ppm))
#   )
```

```{r}
# # Verify if at least one filled value is from the real value fro the same group
# train_soil_filled %>%
#   mutate(site_base = str_remove(site, "\\d+$")) %>%
#   group_by(site_base) %>%
#   summarise(
#     n_site = n(),
#     n_soilk_non_na = sum(!is.na(soilk_ppm)),
#     n_soilp_non_na = sum(!is.na(soilp_ppm))
#   ) %>%
#   arrange(n_soilk_non_na, n_soilp_non_na)
```

```{r}
# # To check which site was missing but now is filled
# train_soil_compare <- train_soil_filled %>%
#   select(site, year, soilk_ppm_old = soilk_ppm, soilp_ppm_old = soilp_ppm) %>%
#   left_join(
#     train_soil_filled2 %>%
#       select(site, year, soilk_ppm_new = soilk_ppm, soilp_ppm_new = soilp_ppm),
#     by = c("site", "year")
#   ) %>%
#   filter(
#     (is.na(soilk_ppm_old) & !is.na(soilk_ppm_new)) |
#     (is.na(soilp_ppm_old) & !is.na(soilp_ppm_new))
#   )
# 
# train_soil_compare
```

### IV. Merge data

#### 1. Check difference

```{r}
train_soil_filled %>% count(site, year) %>% filter(n > 1)
train_meta_clean %>% count(site, year) %>% filter(n > 1)
weather_tain_month_sum_wide %>% count(site, year) %>% filter(n > 1)

setdiff(train_trait_clean$site, weather_tain_month_sum_wide$site)
setdiff(train_trait_clean$site, train_soil_filled$site)
setdiff(train_trait_clean$year, weather_tain_month_sum_wide$year)
setdiff(train_trait_clean$year, train_soil_filled$year)
```

#### 2. Left join

```{r}
train_data <- train_trait_clean %>%
  left_join(weather_tain_month_sum_wide, by = c("site", "year")) %>%
  left_join(train_soil_filled, by = c("site", "year")) %>%
  dplyr::select(-planted_month, -harvested_month, -hybrid)

train_data
write_csv(train_data, "../output/train_data.csv")
```

#### 3. Check merged data

```{r}
# check row numbers
nrow(train_trait_clean)
nrow(train_data)

# check if there is repeatedly join
weather_tain_month_sum_wide %>%
  count(site, year) %>%
  filter(n > 1)

train_soil_clean %>%
  count(site, year) %>%
  filter(n > 1)

# check NA
train_data %>%
  summarise(
    n_weather_na = sum(is.na(mean_dayl.s_stage1)),
    n_soil_na = sum(is.na(soil_ph))
  )
# check which specific sites have problem
train_data %>%
  filter(is.na(soil_ph)) %>%
  distinct(site, year)
```

## Testing data

### Ⅰ. Growing data

#### 1. Set up

##### 1.1 Import taining data

```{r}
test_meta <- read_csv("../data/testing_meta.csv")
test_soil <- read_csv("../data/testing_soil.csv")
test_trait <- read_csv("../data/testing_submission.csv")

#glimpse(test_meta)
#glimpse(test_soil)
#glimpse(test_trait)
```

#### 2. EDA

##### 2.1 Summary

```{r}
summary(test_meta)
summary(test_soil)
summary(test_trait)
```

##### 2.2 Distinct year and site

```{r}
test_meta %>% distinct(year)
test_soil %>% distinct(year)
test_trait %>% distinct(year)

test_meta %>% distinct(site)
test_soil %>% distinct(site)
test_trait %>% distinct(site)

test_trait %>% distinct(hybrid)
```

##### 2.3 Check non-US site

```{r}
# check if all sites are in the US.

# read US boundary
usa <- ne_countries(
  scale = "medium",
  country = "United States of America",
  returnclass = "sf"
)

# convert points to sf objects
points <- test_meta %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# check whether points are within the US
inside <- st_within(points, usa)

# label results (clearer)
test_meta_check <- test_meta %>%
  mutate(in_us = lengths(inside) > 0)

# view points not in the US
not_in_us <- test_meta_check %>%
  filter(!in_us)

not_in_us

```

According to the latitude and longitude, ONH3 is located outside the United States.

##### 2.4 Check A/B and B/A in hybrid

```{r}
hyb_check <- test_trait %>%
  distinct(hybrid) %>%
  separate(hybrid, into = c("p1", "p2"), sep = "/", remove = FALSE) %>%
  mutate(rev = paste(p2, p1, sep = "/"))

sum(hyb_check$rev %in% hyb_check$hybrid)
```

There are 2 pairs that contain A/B and B/A in hybrid.

#### 3. Data cleaning

##### 3.1 Clean test_soil

```{r}
test_soil_clean <- read_csv("../data/testing_soil.csv") %>%
  clean_names() %>%
  rename(soil_ph = soilp_h) %>%
  mutate(
    site = str_remove(site, "_\\d{4}$")
  ) %>%
  group_by(year, site) %>%
  summarise(
    soil_ph   = mean(soil_ph, na.rm = TRUE),
    om_pct    = mean(om_pct, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(c(soil_ph, om_pct, soilk_ppm, soilp_ppm), ~ round(., 2))) %>%
  arrange(desc(year), site)

test_soil_clean

write_csv(test_soil_clean, "../output/test_soil_clean.csv")
test_soil_clean %>% filter(str_detect(site, "\\d{4}$"))
```

##### 3.2 Clean train_trait

```{r}
test_trait_clean <- read_csv("../data/testing_submission.csv") %>%
  clean_names() %>%
  separate(hybrid, into = c("parent1", "parent2"), sep = "/", remove = FALSE) %>%
  dplyr::select(-yield_mg_ha)

test_trait_clean

write_csv(test_trait_clean, "../output/test_trait_clean.csv")
```

##### 3.3 Clean train_meta

```{r}
test_meta_clean <- read_csv("../data/testing_meta.csv") %>%
  clean_names() %>%
  
  mutate(previous_crop = tolower(previous_crop))

test_meta_clean

write_csv(train_meta_clean, "../output/train_meta_clean.csv")
train_meta_clean %>% count(previous_crop, sort = TRUE)
train_meta_clean %>% count(site, year) %>% filter(n > 1)
```

### Ⅱ. Weather data

#### 1. Import data from Daymet

```{r}
# help("download_daymet")

# test_meta_clean_daym <- test_meta_clean %>%
#  mutate(
#    weather = pmap(
#      list(year, site, latitude, longitude),
#     function(.y, .site, .lat, .lon) {
#        download_daymet(
#         site = .site,
#          lat = .lat,
#          lon = .lon,
#          start = .y,
#          end = .y,
#          simplify = TRUE,
#          silent = TRUE
#        ) %>%
#          rename(
#            weather_year = year,
#            weather_site = site,
#            weather_latitude = latitude,
#            weather_longitude = longitude
#          )
#      }
#    )
#  ) %>%
#  unnest(weather) %>%
#  pivot_wider(names_from = measurement,
#              values_from = value
#              ) %>%
#  janitor::clean_names()

# test_meta_clean_daym

# write_csv(test_meta_clean_daym, "../data/test_meta_clean_daym.csv")

weather_test <- read_csv("../data/test_meta_clean_daym.csv")
```

#### 2. EDA

```{r}
# overview
summary(weather_test)

# density plot to see distribution
weather_test %>%
  pivot_longer(cols = dayl_s:vp_pa) %>%
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~name, scales = "free")
```

#### 3. Feature engineering

##### 3.1 Fix grow month

Since planting and harvest date information is not available in the testing dataset, it is difficult to directly align weather data with the actual growing season. To address this, I summarize the range of planting and harvest months for each site using the training data. Specifically, I use the minimum planting month as the start of the growing period and the maximum harvest month as the end. These site-specific month ranges are then used to extract and aggregate weather data for both the training and testing datasets. I assume that this time period reasonably represents the growing season for corn in the testing data.

```{r}
grow_month_test <- grow_month %>%
  bind_rows(
    tibble(
      site = "ONH3",
      start_mon = "Apr",
      end_mon = "Nov"
    )
  )

grow_month_test
```

##### 3.2 Merge grow month with weather data

```{r}

weather_test_grow <- weather_test %>%
  left_join(grow_month_test, by = "site") %>%
  dplyr::select(year, site, previous_crop, start_mon, end_mon, everything())

weather_test_grow

write_csv(weather_test_grow, "../output/weather_test_grow.csv")

setdiff(unique(weather_test_grow$site), unique(grow_month$site))
```

##### 3.3 Create month column for weather info

```{r}
weather_test_month <- weather_test_grow %>%
  # Selecting needed variables
  dplyr::select(year, site, previous_crop, 
                start_mon, end_mon,
                latitude, longitude,
                yday,
                dayl.s = dayl_s, 
                prcp.mm = prcp_mm_day,
                srad.wm2 = srad_w_m_2,
                tmax.c = tmax_deg_c, 
                tmin.c = tmin_deg_c,
                vp.pa = vp_pa
                ) %>%
  # Creating a date class variable
  mutate(date_chr = paste0(year,"/",yday)) %>% 
  mutate(date = as.Date(date_chr, "%Y/%j")) %>%
  # Extracting month from date  
  mutate(month = month(date)) %>%
  mutate(month_abb = month(date, label = T)) %>%
  # calculate GDD
  mutate(
    tmean = (tmax.c + tmin.c) / 2,
    gdd = pmax(tmean - 10, 0) # T-base: 10 for corn https://fieldreport.caes.uga.edu/publications/C1320/growing-degree-days-for-sweet-corn-in-southern-georgia/
    )

weather_test_month
```

##### 3.4 Cut growing season into 4 stages

```{r}
month_levels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

weather_stage_test <- weather_test_month %>%
  mutate(
    month_abb = factor(as.character(month_abb), levels = month_levels, ordered = TRUE),
    start_mon = factor(start_mon, levels = month_levels, ordered = TRUE),
    end_mon   = factor(end_mon, levels = month_levels, ordered = TRUE)
  ) %>%
  filter(month_abb >= start_mon & month_abb <= end_mon) %>%
  group_by(year, site) %>%
  mutate(stage = ntile(yday, 4)) %>%
  ungroup() 

weather_stage_test
```

##### 3.5 Summarize daily weather variables based on growing stage

```{r}
weather_test_month_sum <- weather_stage_test %>%
    mutate(
    month_abb = factor(as.character(month_abb), levels = month_levels, ordered = TRUE),
    start_mon = factor(start_mon, levels = month_levels, ordered = TRUE),
    end_mon   = factor(end_mon, levels = month_levels, ordered = TRUE)
  ) %>%
  filter(month_abb >= start_mon & month_abb <= end_mon) %>% 
  group_by(year, site, previous_crop, stage) %>%
  summarise(across(.cols = c(dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa
                             ),
                   .fns = mean,
                   .names = "mean_{.col}"
                   ),
            across(.cols = c(prcp.mm, gdd),
                   .fns = sum,
                   .names = "sum_{.col}"
                   )
            ) %>%
  ungroup()

weather_test_month_sum
```

##### 3.6 Verify the calculation

```{r}
weather_stage_test %>%
  filter(year == 2024 & 
           site == "DEH1" &
           previous_crop == "soybean" &
           stage == 1) %>%
  summarise(tmax.c = mean(tmax.c),
            prcp.mm = sum(prcp.mm))
```

##### 3.7 Include stage in the column name

```{r}
weather_test_month_sum_wide <- weather_test_month_sum %>%
  pivot_longer(mean_dayl.s:sum_gdd) %>%
  mutate(varname = paste0(name, "_stage",stage)) %>%
  dplyr::select(-name, -stage) %>%
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(where(is.numeric), 
                ~round(.,2)
                )
         ) 

weather_test_month_sum_wide

write_csv(weather_test_month_sum_wide, "../output/weather_test_month_sum_wide.csv")
```

### Ⅲ. Soil data

#### 1. Create full soil info

```{r}
# check which site and year only in test_trait not in test_meta
test_trait_clean %>% 
  anti_join(test_meta_clean, by = c("year", "site"))

# check if there is extra info in meta
test_meta_clean %>% 
  anti_join(test_trait_clean, by = c("year", "site"))
```

```{r}
test_soil_full <- test_meta_clean %>%
  left_join(
    test_soil_clean,
    by = c("year", "site")
  ) %>%
   dplyr::select(-previous_crop)

test_soil_full
```

#### 2. Check the missing info

```{r}
# area
test_soil_full %>%
  summarise(
    om_pct_na = sum(is.na(om_pct)),
    soil_ph_na = sum(is.na(soil_ph)),
    soilk_ppm_na = sum(is.na(soilk_ppm)),
    soilp_ppm_na = sum(is.na(soilp_ppm))
  )

# line
test_soil_full %>%
  filter(
    is.na(om_pct) |
    is.na(soil_ph) |
    is.na(soilk_ppm) |
    is.na(soilp_ppm)
  )
```

#### 3. Keep only the site-years that were pulled

```{r}
test_soil_needpull <- test_soil_full %>%
  filter(
    is.na(om_pct) |
    is.na(soil_ph) |
    is.na(soilk_ppm) |
    is.na(soilp_ppm)
  ) %>% distinct(site, latitude, longitude)

test_soil_needpull
```

#### 4. Pull data from SoilGrid

```{r}
glimpse(test_soil_needpull)

summary(test_soil_needpull$latitude)
summary(test_soil_needpull$longitude)

test_soil_needpull %>% 
  filter(is.na(latitude) | is.na(longitude))
```

```{r}
# SoilGrids pull
# sg_input <- test_soil_needpull %>%
#   filter(!is.na(latitude), !is.na(longitude)) %>%
#   distinct(latitude, longitude) %>%
#   mutate(id = paste0("pt_", row_number())) %>%
#   select(id, latitude, longitude)
# 
# soilgrid_raw <- fetchSoilGrids(
#   x = sg_input,
#   loc.names = c("id", "latitude", "longitude"),
#   variables = c("phh2o", "soc"),
#   depth_intervals = "0-5"
# )

```

#### 5. Organize data

```{r}
# names(soilgrid_raw)
```

```{r}
# h <- horizons(soilgrid_raw) %>% as.data.frame()
# s <- site(soilgrid_raw) %>% as.data.frame()
# 
# names(s)
# names(h)
```

```{r}
# soilgrid_fill_test  <- s %>%
#   left_join(
#     h %>%
#       select(id, phh2omean, socmean),
#     by = "id"
#   ) %>%
#   transmute(
#     latitude,
#     longitude,
#     soil_ph_sg = phh2omean,
#     om_pct_sg = socmean * 0.1724
#   )
# 
# soilgrid_fill_test 
# 
# write_csv(soilgrid_fill, "../output/soilgrid_fill_test.csv")
```

#### 6. Merge back

```{r}

soilgrid_fill_test <- read_csv("../output/soilgrid_fill_test.csv")

test_soil_filled <- test_soil_full %>%
  left_join(soilgrid_fill_test , by = c("latitude", "longitude")) %>%
  mutate(
    soil_ph = coalesce(soil_ph, soil_ph_sg),
    om_pct  = coalesce(om_pct, om_pct_sg)
  ) %>%
  select(-soil_ph_sg, -om_pct_sg, -longitude, -latitude)

test_soil_filled
```

```{r}
# check NAs
test_soil_filled %>%
  summarise(
    soil_ph_na = sum(is.na(soil_ph)),
    om_pct_na  = sum(is.na(om_pct)),
    soilk_ppm_na = sum(is.na(soilk_ppm)),
    soilp_ppm_na = sum(is.na(soilp_ppm))
  )
```

```{r}
# label ones with SoilGrid data

# test_soil_filled <- test_soil_full %>%
#   left_join(soilgrid_fill, by = c("latitude", "longitude")) %>%
#   mutate(
#     soil_ph_from_sg = as.integer(is.na(soil_ph) & !is.na(soil_ph_sg)),
#     om_pct_from_sg  = as.integer(is.na(om_pct)  & !is.na(om_pct_sg)),
#     soil_ph = coalesce(soil_ph, soil_ph_sg),
#     om_pct  = coalesce(om_pct, om_pct_sg)
#   ) %>%
#   select(-soil_ph_sg, -om_pct_sg)
# 
# test_soil_filled
```

```{r}
# test_soil_filled2 <- test_soil_filled %>%
#   mutate(
#     site_base = str_remove(site, "\\d+$")
#   ) %>%
#   group_by(site_base) %>%
#   mutate(
#     soilk_base_mean = mean(soilk_ppm, na.rm = TRUE),
#     soilp_base_mean = mean(soilp_ppm, na.rm = TRUE)
#   ) %>%
#   ungroup() %>%
#   mutate(
#     soilk_ppm = if_else(
#       is.na(soilk_ppm) & is.finite(soilk_base_mean),
#       soilk_base_mean,
#       soilk_ppm
#     ),
#     soilp_ppm = if_else(
#       is.na(soilp_ppm) & is.finite(soilp_base_mean),
#       soilp_base_mean,
#       soilp_ppm
#     )
#   ) %>%
#   select(-site_base, -soilk_base_mean, -soilp_base_mean)
# 
# test_soil_filled2
```

```{r}
# test_soil_filled2 %>%
#   summarise(
#     soilk_ppm_na = sum(is.na(soilk_ppm)),
#     soilp_ppm_na = sum(is.na(soilp_ppm))
#   )
```

```{r}
# test_soil_filled %>%
#   mutate(site_base = str_remove(site, "\\d+$")) %>%
#   group_by(site_base) %>%
#   summarise(
#     n_site = n(),
#     n_soilk_non_na = sum(!is.na(soilk_ppm)),
#     n_soilp_non_na = sum(!is.na(soilp_ppm))
#   ) %>%
#   arrange(n_soilk_non_na, n_soilp_non_na)
```

```{r}
# test_soil_compare <- test_soil_filled %>%
#   select(site, year, soilk_ppm_old = soilk_ppm, soilp_ppm_old = soilp_ppm) %>%
#   left_join(
#     test_soil_filled2 %>%
#       select(site, year, soilk_ppm_new = soilk_ppm, soilp_ppm_new = soilp_ppm),
#     by = c("site", "year")
#   ) %>%
#   filter(
#     (is.na(soilk_ppm_old) & !is.na(soilk_ppm_new)) |
#     (is.na(soilp_ppm_old) & !is.na(soilp_ppm_new))
#   )
# 
# train_soil_compare
```

### IV. Merge data

#### 1. Check difference

```{r}
test_soil_filled %>% count(site, year) %>% filter(n > 1)
test_meta_clean %>% count(site, year) %>% filter(n > 1)
weather_test_month_sum_wide %>% count(site, year) %>% filter(n > 1)

setdiff(test_trait_clean$site, weather_test_month_sum_wide$site)
setdiff(test_trait_clean$site, test_soil_filled$site)
setdiff(test_trait_clean$year, weather_test_month_sum_wide$year)
setdiff(test_trait_clean$year, test_soil_filled$year)
```

#### 2. Left join

```{r}
test_data <- test_trait_clean %>%
  left_join(weather_test_month_sum_wide, by = c("site", "year")) %>%
  left_join(test_soil_filled, by = c("site", "year")) %>%
  dplyr::select(-hybrid)

test_data
write_csv(test_data, "../output/test_data.csv")
```

#### 3. Check merged data

```{r}
# check row numbers
nrow(test_trait_clean)
nrow(test_data)

# check if there is repeatedly join
weather_test_month_sum_wide %>%
  count(site, year) %>%
  filter(n > 1)

test_soil_clean %>%
  count(site, year) %>%
  filter(n > 1)

# check NA
test_data %>%
  summarise(
    n_weather_na = sum(is.na(mean_dayl.s_stage1)),
    n_soil_na = sum(is.na(soil_ph))
  )
# Check which specific site has problem
test_data %>%
  filter(is.na(soil_ph)) %>%
  distinct(site, year)
```
