---
title: "XGboost_Vinicius-version"
format: html
editor: visual
---

# OnDemand packages

```{r}
# class_packages <- c("tidymodels","tidyverse","vip","ranger","finetune", "parnship", "reticulate", "xgboost", "kernlab",doParallel", "lme4", "here", "lmer4", "noplr")

shared_lib_path <- "/home/vfa43171/crss8030/vini/shared_R_ribs"

#class_packages <- c("glmnet","lme4")

.libPaths(c(shared_lib_path, .libPaths()))

```

```{r}
.libPaths(c("/home/vfa43171/crss8030/vini/shared_R_ribs", .libPaths()))
```

```{r Library}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
library(caret)       # Other great library for Machine Learning 
library(glmnet)
```

# Load the data set

```{r}
corn_data <- read_csv("/home/vfa43171/2026dsa_finalproject_group11/data/weather_final/trainingdata_vinicius.csv")
summary(corn_data)
```

# ML workflow

## 1. Pre-processing

```{r}
set.seed(2026)
weather_split <- initial_split(
  corn_data,
  prop = 0.7,
  strata = yield_mg_ha_mean
)
weather_split
```

### a. Data split

For a data split, I will use **70% training / 30 % testing**

```{r weather_train}
corn_train <- training(weather_split) #70% of data
corn_train
```

```{r}
corn_test <- testing(weather_split)
corn_test
```

### b. Distribution of target variable

```{r}
ggplot() +
  geom_density(data = corn_train,
               aes(x = yield_mg_ha_mean),
               color = "red") +
  geom_density(data = corn_test,
               aes(x = yield_mg_ha_mean),
               color = "blue")
```

### c. Data processing with recipe

```{r corn_recipe}
corn_recipe <- recipe(yield_mg_ha_mean ~ ., data = corn_train) %>%
  step_rm(year, site, hybrid, previous_crop) #It was getting lots of unknown/NAs
corn_recipe
```

```{r weather_prep}
corn_prep <- corn_recipe %>%
  prep()

corn_prep
```

## 2. Training

### a. Model specification

```{r xgb_spec}
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  min_n = tune(),
  learn_rate = tune()
  ) %>%
  
  set_engine("xgboost") %>%
  set_mode("regression") 

xgb_spec
```

### b. Cross-validation setup

```{r}
set.seed(2351)
resampling_foldcv <- vfold_cv(corn_train, v=10) 

resampling_foldcv
resampling_foldcv$splits[[1]]
```

### c. Hyperparameter grid with Latin Hypercube Sampling

```{r}
xgb_grid <- grid_latin_hypercube(
  tree_depth(), 
  min_n(), 
  learn_rate(), 
  trees(), 
  size = 20) #It should be n=100, when running on Demand 

xgb_grid
```

```{r}
ggplot(data = xgb_grid,
       aes(x = tree_depth,
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate), 
                 size = trees),
             alpha = 0.5,
             show.legend = F
             )
```

## 3. Model Tuning

```{r}
set.seed(2026)
registerDoParallel(cores = parallel::detectCores() -1)

xgb_res <- tune_race_anova(object = xgb_spec,
                           preprocessor = corn_recipe,
                           resamples = resampling_foldcv,
                           grid = xgb_grid,
                           control = control_race(save_pred = T))
stopImplicitCluster()

beepr::beep()
xgb_res$.metric[[2]]
```

```{r}
plot_race(xgb_res)
```

## 4. Select Best Models

We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:

```{r}
# Based on lowest RMSE
best_rmse <- xgb_res %>%
  select_best(metric = "rmse") %>%
  mutate(source = "best_rmse")

best_rmse
```

```{r}
# Based on lowers RMSE within 1% loss
best_rmse_pct_loss <- xgb_res %>%
  select_by_pct_loss("min_n",
                     metric = "rmse",
                     limit = 1) %>%
  mutate(source = "best_rmse_pct_loss")

best_rmse_pct_loss
```

```{r}
# Based on lowest RMSE within 1 se
best_rmse_one_std_err <- xgb_res %>% 
  select_by_one_std_err(metric = "rmse",
                        eval_time = 100,
                        trees
                        )%>% 
  mutate(source = "best_rmse_one_std_err")

best_rmse_one_std_err
```

```{r}
# Based on greatest R2
best_r2 <- xgb_res %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```

```{r}
# Based on lowers R2 within 1% loss
best_r2_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rsq",
                     limit = 1
                     ) %>% 
  mutate(source = "best_r2_pct_loss")

best_r2_pct_loss
```

```{r}
# Based on lowest R2 within 1 se
best_r2_one_std_error <- xgb_res %>% 
  select_by_one_std_err(metric = "rsq",
                        eval_time = 100,
                        trees
                        ) %>%
  mutate(source = "best_r2_one_std_error")

best_r2_one_std_error
```

## Compare and finalize model

```{r}
best_rmse %>% 
  bind_rows(best_rmse_pct_loss, 
            best_rmse_one_std_err, 
            best_r2, 
            best_r2_pct_loss, 
            best_r2_one_std_error)
```

## 5. Final Specification

```{r}
final_spec <- boost_tree(
  trees = best_r2$trees,           # Number of boosting rounds (trees)
  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
  min_n = best_r2$min_n,           # Minimum number of samples to split a node
  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```

```{r Model Workflow}
final_workflow <- workflow() %>%
  add_recipe(corn_recipe) %>%
  add_model(final_spec) %>%
  finalize_workflow(best_rmse)
```

## 6. Final Fit and Predictions

## Validation

```{r final_fit}
set.seed(10)
final_fit <- last_fit(final_spec,
                corn_recipe,
                split = weather_split)

final_fit %>%
  collect_predictions()
```

## 7. Evaluate on Test Set

```{r}
final_fit %>%
  collect_metrics()
```

```{r}
test_met <- final_fit %>%
  collect_metrics() %>% 
  mutate(estimate = round(.estimate, 3))
```

## 8. Evaluate on Training Set

```{r}
final_spec %>%
  fit(yield_mg_ha_mean ~ .,
      data = bake(corn_prep, 
                  corn_train)) %>%
  augment(new_data = bake(corn_prep, 
                          corn_train)) %>% 
  rmse(yield_mg_ha_mean, .pred) %>%
  bind_rows(
    
    
# R2
final_spec %>%
  fit(yield_mg_ha_mean ~ .,
      data = bake(corn_prep, 
                  corn_train)) %>%
  augment(new_data = bake(corn_prep, 
                          corn_train)) %>% 
  rsq(yield_mg_ha_mean, .pred))
```

## 9. Predicted vs Observed Plot

```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha_mean,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(0, 40)) +
  scale_y_continuous(limits = c(0, 40)) 
```

```{r Publication Ready}
publication_ready <- final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha_mean,
             y = .pred)) +

  geom_point(aes(fill = yield_mg_ha_mean), shape = 21, alpha = 0.7, show.legend = F) +
  
  scale_fill_viridis_c(option = "H") +
  
  geom_abline(color = "red", linetype = 2) +
  
  geom_smooth(method = "lm") +
  labs(x = "Observed Yield (Mg/Ha)",
       y = "Predicted Yield (Mg/Ha)")+
  
  annotate("label", x = Inf, y = -Inf,
           label = paste0(
             "R²: ", test_met$estimate[test_met$.metric == "rsq"],
             "\nRMSE: ", test_met$estimate[test_met$.metric == "rmse"]),
           hjust = 1, vjust = -0.5) +
  
  scale_x_continuous(limits = c(0, 40)) +
  scale_y_continuous(limits = c(0, 40)) +
  
  # Changing the theme of the plot
    theme(panel.background = element_rect(fill = "gray82"))

ggsave(plot = publication_ready, 
       path = ("../output/"),
       filename = "publication_ready_graph.png",
       height = 6,
       width = 9,
       dpi = 600)
```

## 10. Variable Importance

```{r}
final_spec %>%
  fit(yield_mg_ha_mean ~ .,
         data = bake(corn_prep, corn_train)) %>% #There little change in variable importance if you use full dataset
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)
```

## 11. Applying on 2024 test set

### a. Loading test set

```{r}
weather_features <- read.csv("../data/ML_data_Vinicius/testingdata_vinicius.csv") 
```

```{r}
test_submission <- read.csv("../data/testing/testing_submission.csv") 
```

```{r}
final_workflow_model <- final_workflow %>%
  fit(data = corn_train)
```

### b. Prediction

```{r}
predict_2024 <- predict(final_workflow_model, new_data = weather_features) %>%
bind_cols(weather_features %>% select(site, year, hybrid)) %>%
  rename(yield_mg_ha = .pred)
```

```{r}
final_answer <- test_submission %>%
  mutate(yield_mg_ha = predict_2024$yield_mg_ha)
```

### c. Exporting CSV file

```{r}
write_csv(final_answer, "../output/Yield_Predictions_2024.csv")
```

```{r Preparing to run onDemand}
#knitr::purl("XGBoost_Vinicius.qmd", output = "sapelo_scriptXGB.R", documentation = 0)
```
